---
title: "marchmadness2017"
output:
  pdf_document: default
  html_document: default
---
 
# Lotter Picks
UCLA (No. 1 Lonzo Ball), Kansas (No. 2 Josh Jackson), Duke (No. 3 Jayson Tatum and No. 9 Harry Giles), Kentucky (No. 4 De'Aaron Fox and No. 5 Malik Monk), Arizona (No. 6 Lauri Markkanen), Florida State (No. 7 Jonathan Issac) and Gonzaga (No. 10 Zach Collins).
http://www.espn.com/blog/statsinfo/post/_/id/130371/using-recent-history-to-pick-a-national-champion
 
# Source Data
 
```{r}
library(plyr)
library(dplyr)
library(stringr)
library(PlayerRatings)
library(car)
library(randomForest)
library(tidyr)
# library(gbm)

sourcepath <- "C:/Users/Amy/Documents/GitHub/mm2017/scripts/"

reg_det<-read.csv(paste0(inpath, "RegularSeasonDetailedResults.csv"), stringsAsFactors = FALSE, header = TRUE)
#tourney.details<-read.csv(paste0(inpath, "TourneyDetailedResults.csv"), stringsAsFactors = FALSE, header = TRUE)

# 32 seconds to do all processing
#source(file = paste0(sourcepath,"proc_data.R"))

```

## Munge: End of Season Stats

```{r}
# stack these
reg_det_win <- select(reg_det, -Lfgm, -Lfga, -Lfgm3, -Lfga3, -Lftm, -Lfta, -Lor, -Ldr, -Last, -Lto, -Lstl, -Lblk, -Lpf, -Lteam, -Lscore, -Wloc)
## rename
names(reg_det_win) <- str_replace(names(reg_det_win), pattern = "^W", "")

reg_det_lose <- select(reg_det, -Wfgm, -Wfga, -Wfgm3, -Wfga3, -Wftm, -Wfta, -Wor, -Wdr, -Wast, -Wto, -Wstl, -Wblk, -Wpf, -Wteam, -Wscore, -Wloc)

names(reg_det_lose) <- str_replace(names(reg_det_lose), pattern = "^L", "")

out_reg_det <- rbind(reg_det_win, reg_det_lose)

# split data set into wins and losses...
reg_det_win$win <- "win"
reg_det_lose$win <- "loss"

# ...then recombine to model: half of rows represent a win, half represent a loss
mod_out_reg_det <- rbind(reg_det_win, reg_det_lose)
mod_out_reg_det$win <- as.factor(mod_out_reg_det$win) # convert to factor for randomForest

start <- Sys.time()
m_rf1 <- randomForest(win ~ fgm + fga + fgm3 + fga3 + ftm + fta + or + dr + ast + to + stl + blk + pf, data = mod_out_reg_det)
end <- Sys.time()

varImpPlot(m_rf1)
```

## Munge: Data for Logistic Model 

* 

```{r}

```

## Summarize: average stats per team for end of season

```{r}
## summarize the stats
proc_reg_det <- group_by(out_reg_det, Season, team) %>%
  arrange(Daynum) %>% 
  dplyr::summarise(fgm=mean(fgm),
                   fga=mean(fga),
                   fg_pct=(fgm/fga),
                   fgm3=mean(fgm3),
                   fga3=mean(fga3),
                   fg3_pct=(fgm3/fga3),
                   ftm=mean(ftm),
                   fta=mean(fta),
                   ft_pct=(ftm/fta),
                   or=mean(or),
                   dr=mean(dr),
                   ast=mean(ast),
                   to=mean(to),
                   stl=mean(stl),
                   blk=mean(blk),
                   pf=mean(pf))
```

# Model: Logistic Regression of Regular season outcomes

* http://multithreaded.stitchfix.com/blog/2016/11/22/nba-rankings/

```{r}

```
